Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-03-26 WSSSPE1_report/WSSSPE1_summary_report.tex
minor fixes in communities

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item * James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

Drawing on experiences from high-energy physics,~\cite{Vay_WSSSPE} proposed
developing teams of technical specialists able to overcome a lack of
coordination between projects.  Maximization of scientific output requires
maximizing the usability of the scientific codes while minimizing the cost of
developing and supporting those codes.  This included targeting different
architectures for their software to be deployed, as well as coordination
between technically-focused individuals and usage of a common scripting
language between projects.  Instead of fragmenting the development of
simulation codes across institutions, the paper suggests that a cohesive
strategy reducing duplication and increasing coordination will broadly increase
the efficiency across institutions.  The approach proposed is of de-fragmenting
the existing ecosystem in a non-disruptive way.

@@

In~\cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
source software development at Kitware.  In particular, they focus on their
success at growing their community of users through enabling multiple channels
of communication, directly reaching out to individuals, and lowering the
barrier to entry for contributions.  This involves providing clear,
test-oriented and review-based mechanisms for evaluating contributions,
permissive licenses, and a service-based model for sustaining development.
This model enables Kitware to review both public funding, as well as private
funding to support improvements and targeted developments of the software.

At NESCent, a combination of in-house informatics individuals and domain
scientists collaborate to develop software to study evolutionary science.  The
report~\cite{Cranston_WSSSPE} studied the success of a ``hackathon'' model
for development, where short-form, hands-on events combining users,
researcher-developers and software engineers targeted specific code
improvements.  From this experiment, the authors identified sweveral key
outcomes as well as lessons-learned -- specifically, the co-localization of
developers was seen as having a strong impact, enabling casual conversation
that led to discrete outcomes.  The formation of the discussion mailing list,
specifically in response to the social capital built at the hackathon, was seen
as spurring on longer-term benefits to the community and fostering
sustainability.

\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
collaborations supporting tool development for Open Science.  This software
collective, organized around the statistical programming environment R,
develops access mechanisms for data repositories and attempts to reduce the
barrier to entry for individuals wanting to access data repositories and study
the data contained therein.  The collective fosters direct collaboration
between individuals and data providers, designed to ``train academics in
reproducible science workflows focused around R.''  The two central challenges
to this goal were identified as engagement of existing users within ecology and
evolutionary biology (EEB), and how can the community make inroads and traction
in other disciplines?  Currently, the collective is exploring addressing these
challenges through use of social media and holding workshops and hackathons.
This helps to both raise the profile of the collective within EEB and in other
domains.  However, the overarching challenge identified in the paper was that
of incentivizing maintenance of software, which is difficult in academia.

@@

The Christopherson paper outlines the degree to which research relies on high
quality software. There are often barriers and a lack of suitable incentives
for researchers to embrace software engineering principals. The Water Science
Software Institute is working to lower some of these barriers through an Open
Community Engagement Process. This is a 4-step iterative development process
that incorporates Agile development principals.

@@

The Christopherson paper reports on the application of Steps 1-3 to a
computational modeling framework developed in the 1990s. Step 1 was realized as
a 2-day, in person specifications meeting and code walkthrough. Step 2 was a
5-day hackathon to develop working code, and Step 3 a 3-day hackathon to refine
the code based on new requirements. The team worked on small, low-risk units of
code. It was challenging, revealed unanticipated obstacles, programmers had to
work together, and experimentation was encouraged.

@@

The Pierce paper described how science gateways can provide a user-friendly
entry to complex cyberinfrastructure. The paper opens with a description of the
explosion of use of just a few gateways. Over 3.5 years, more than 7,000
biologists have run phylogenetic codes on supercomputers using the CIPRES
Science Gateway. In 1 year, over 120 scientists from 50 institutions used the
UltraScan Science Gateway, increasing the sophistication of analytical
ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG)
registered 100+ users who used 250,000 CPU hours in only a few months.

@@

The Zentner paper describes experiences and challenges with the large
nanoHUB.org community. The authors define community as a ``body of persons of
common and especially professional interests scattered through a larger
society.'' This makes support challenging because of the diversity of
viewpoints and needs. The group constantly examines its policies to determine
whether they are indirectly alienating part of the community or encouraging
particular types of use.

@@

The Terrel paper describes support for the Python scientific community through
two major efforts - the SciPy conference and the NumFOCUS foundation. Reliance
on software in science has driven a huge demand for development, but this
development is typically done as a side effort and often in a rush to publish
without documentation and testing. While created by academics, software support
often falls to industrial institutions. SciPy brings together industry,
government, and academics to share their code and experience in an open
environment.

@@

The L\"{o}ffler paper describes the Cactus project which was started in 1996 by
participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh
and thorns, core and module, model - a community-oriented framework that allows
researchers to easily work together with reusable and extendable software
elements. Modules are compiled into an executable and can remain dormant,
becoming active only when parameters and simulation data dictate. Users can
use modules written by others or can write their own modules without changing
other code. The community has grown and diversified beyond the original science
areas.

@@

The Wilkins-Diehr paper describes an NSF software institute effort to build a
community of those creating science gateways or web portals for science. These
gateways, as described in some of the other papers in this section, can be
quite capable, having a remarkable impact on some parts of science.

>>>>

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item * James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item * Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

Drawing on experiences from high-energy physics, \cite{Vay_WSSSPE} proposed
developing teams of technical specialists able to overcome a lack of
coordination between projects.  Maximization of scientific output requires
maximizing the usability of the scientific codes while minimizing the cost of
developing and supporting those codes.  This included targeting different
architectures for their software to be deployed, as well as coordination
between technically-focused individuals and usage of a common scripting
language between projects.  Instead of fragmenting the development of
simulation codes across institutions, the paper suggests that a cohesive
strategy reducing duplication and increasing coordination will broadly increase
the efficiency across institutions.  The approach proposed is of de-fragmenting
the existing ecosystem in a non-disruptive way.

@@

In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
source software development at Kitware.  In particular, they focus on their
success at growing their community of users through enabling multiple channels
of communication, directly reaching out to individuals, and lowering the
barrier to entry for contributions.  This involves providing clear,
test-oriented and review-based mechanisms for evaluating contributions,
permissive licenses, and a service-based model for sustaining development.
This model enables Kitware to review both public funding, as well as private
funding to support improvements and targeted developments of the software.

At NESCent, a combination of in-house informatics individuals and domain
scientists collaborate to develop software to study evolutionary science.  The
report~\cite{Cranston_WSSSPE} studied the success of a ``hackathon'' model
for development, where short-form, hands-on events combining users,
researcher-developers and software engineers targeted specific code
improvements.  From this experiment, the authors identified several key
outcomes as well as lessons-learned -- specifically, the co-localization of
developers was seen as having a strong impact, enabling casual conversation
that led to discrete outcomes.  The formation of the discussion mailing list,
specifically in response to the social capital built at the hackathon, was seen
as spurring on longer-term benefits to the community and fostering
sustainability.

\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
collaborations supporting tool development for Open Science.  This software
collective, organized around the statistical programming environment R,
develops access mechanisms for data repositories and attempts to reduce the
barrier to entry for individuals wanting to access data repositories and study
the data contained therein.  The collective fosters direct collaboration
between individuals and data providers, designed to ``train academics in
reproducible science workflows focused around R.''  The two central challenges
to this goal were identified as engagement of existing users within ecology and
evolutionary biology (EEB), and how the community could make inroads and traction
in other disciplines.  Currently, the collective is exploring addressing these
challenges through use of social media and holding workshops and hackathons.
This helps to both raise the profile of the collective within EEB and in other
domains.  However, the overarching challenge identified in the paper was that
of incentivizing maintenance of software, which is difficult in academia.

@@

Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree to which research relies on high
quality software. There are often barriers and a lack of suitable incentives
for researchers to embrace software engineering principals. The Water Science
Software Institute is working to lower some of these barriers through an Open
Community Engagement Process. This is a 4-step iterative development process
that incorporates Agile development principals.

@@

Christopherson reports on the application of Steps 1-3 to a
computational modeling framework developed in the 1990s. Step 1 was realized as
a 2-day, in person specifications meeting and code walkthrough. Step 2 was a
5-day hackathon to develop working code, and Step 3 was a 3-day hackathon to refine
the code based on new requirements. The team worked on small, low-risk units of
code. It was challenging, revealed unanticipated obstacles, programmers had to
work together, and experimentation was encouraged.

@@

Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can provide a user-friendly
entry to complex cyberinfrastructure. The paper opens with a description of the
explosion of use of just a few gateways. Over 3.5 years, more than 7,000
biologists have run phylogenetic codes on supercomputers using the CIPRES
Science Gateway. In 1 year, over 120 scientists from 50 institutions used the
UltraScan Science Gateway, increasing the sophistication of analytical
ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG)
registered 100+ users who used 250,000 CPU hours in only a few months.

@@

Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and challenges with the large
nanoHUB.org community. The authors define community as a ``body of persons of
common and especially professional interests scattered through a larger
society.'' This makes support challenging because of the diversity of
viewpoints and needs. The group constantly examines its policies to determine
whether they are indirectly alienating part of the community or encouraging
particular types of use.

@@

Terrel~\cite{Terrel_WSSSPE} describes support for the Python scientific community through
two major efforts - the SciPy conference and the NumFOCUS foundation. Reliance
on software in science has driven a huge demand for development, but this
development is typically done as a side effort and often in a rush to publish
without documentation and testing. While created by academics, software support
often falls to industrial institutions. SciPy brings together industry,
government, and academics to share their code and experience in an open
environment.

@@

L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project which was started in 1996 by
participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh
and thorns, core and module, model - a community-oriented framework that allows
researchers to easily work together with reusable and extendable software
elements. Modules are compiled into an executable and can remain dormant,
becoming active only when parameters and simulation data dictate. Users can
use modules written by others or can write their own modules without changing
other code. The community has grown and diversified beyond the original science
areas.

@@

Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF software institute effort to build a
community of those creating science gateways or web portals for science. These
gateways, as described in some of the other papers in this section, can be
quite capable, having a remarkable impact on some parts of science.
